Coordinated monitoring of the Piney Point wastewater discharge into Tampa Bay: Data synthesis and reporting

### 2022-05-23 manu-draft.Rmd
change font to 11, check spelling

The release of wastewater from Piney Point in the spring of 2021 was an unfortunate event that mobilized dozens of local resource managers and stakeholders to track the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  Area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for these applications.  More holistically assessing the value of open science tools is also needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts.

From a data science perspective, several lessons were learned from the need to rapidly synthesize multiple sources of information collected during the response-based monitoring.  Broader use of "tidy" data principles would significantly decrease the processing time required to create data in a usable format for analysis.  Data entry sheets or digital platforms for sharing data should be structured to more easily transform the the results into tidy format (i.e., one observation per row, one varialbe in a column, one value per cell).  However, this requires coordination among multiple entities as to which standards are acceptable for use in digital information storage and even additional training activities to bridge communication gaps between field staff and analysts. Data wrangling will always be a component of analysis, but the time spent doing so could be reduced by communication between institutions to adopt similar data standards.  Finding solutions for sharing data was also problematic.  Although Google Drive was useful, not everyone was able to access the information given IT limitations that were agency-specific.  Moreover, hosting data on Google Drive was contingent on trust and willingness to share data with TBEP.  The TBEP did not have access to all data collected for Piney Point, despite the dozens of partners that did share data with the organization.  Building and sustaining partnerships is a fundamental mission of the National Estuary Program.  Past successes of TBEP have helped build this trust and the continued success of the organization is dependent on sustaining and growing partnerships in the future.

@@

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive, data were synthesized using R-based tools, data and dashboard code were uploaded in a unified format on GitHub that included automated testing, and then the app was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

@@

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st 2021 through September.  Values show the total number of unique sites samples for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total for each monitoring type.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```

```{r demtab}
cap.val <- 'Top ten locations of users by city, state, and country for the Piney Point dashboard.  Numbers show the total users and percentages from the total for all locations.  Total cities, states, and countries that accessed the dashboard are shown in the column headers in parentheses.  Summaries are from May 13th, 2021 when analytics were included in the dashboard to October 1st, 2021.'
knitr::kable(demtab, booktabs = T, caption = cap.val)
```

>>>>

The release of wastewater from Piney Point in the spring of 2021 mobilized dozens of local resource managers and stakeholders to track the potential impacts to Tampa Bay. The communication tools and workflows used to synthesize the data collected in this effort were useful for informing the management response and the analytics of our dashboard suggested that the products created by TBEP were valuable resources in 2021.  However, dashboard analytics are not the only metric of success and additional information can help assess the effectiveness of these tools to efficiently and transparently communicate science for decision-making.  Area researchers and managers were able to use screenshots from the dashboard to communicate results in real time, although there was no formal metric for these applications.  More holistically assessing the value of open science tools is also needed, such as how they may be used to inform policy or regulatory changes aimed at minimizing environmental impacts.

From a data science perspective, several lessons were learned from the need to rapidly synthesize multiple sources of information collected during the response-based monitoring.  Broader use of "tidy" data principles would significantly decrease the processing time required to create data in a usable format for analysis.  Data entry sheets or digital platforms for sharing data should be structured to more easily transform the the results into tidy format (i.e., one observation per row, one variable in a column, one value per cell).  However, this requires coordination among multiple entities as to which standards are acceptable for use in digital information storage and even additional training activities to bridge communication gaps between field staff and analysts. Data wrangling will always be a component of analysis, but the time spent doing so could be reduced by communication between institutions to adopt similar data standards.  Finding solutions for sharing data was also problematic.  Although Google Drive was useful, not everyone was able to access the information given IT limitations that were agency-specific.  Moreover, hosting data on Google Drive was contingent on trust and willingness to share data with TBEP.  The TBEP did not have access to all data collected for Piney Point, despite the dozens of partners that did share data with the organization.  Building and sustaining partnerships is a fundamental mission of the National Estuary Program.  Past successes of TBEP have helped build this trust and the continued success of the organization is dependent on sustaining and growing partnerships in the future.

@@

```{r synthesis, fig.cap = 'Data synthesis workflow for Piney Point data.  Raw data were uploaded by partners to Google Drive, data were synthesized using R-based tools, data and dashboard code were uploaded in a unified format on GitHub that included automated testing, and the app was deployed using Shiny Server.'}
knitr::include_graphics(here('figs/synthesis.JPG'))
```

@@

```{r efftab}
cap.val <- 'Sampling effort for response-based monitoring by partners from April 1st 2021 through September.  Values show the total number of unique sites sampled for water quality, algae (phytoplankton), and seagrass/macroalgae in Tampa Bay.  Values in parentheses show the percentage of the total for each monitoring type.'
knitr::kable(efftab, booktabs = T, caption = cap.val)
```

```{r demtab}
cap.val <- 'Top ten locations of users by city, state, and country for the Piney Point dashboard.  Numbers show the total users and percentages from the total for all locations.  Total cities, states, and countries that accessed the dashboard are shown in the column headers in parentheses.  Summaries are from May 13th, 2021 when analytics were included in the dashboard through September.'
knitr::kable(demtab, booktabs = T, caption = cap.val)
```
